---
title: "Intro to Data Viz"
format: html
---

## Set Up

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(janitor)
library(scales)
library(leaflet)
library(DT)
```

## read Data

```{r}
escape_raw <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e")
```

## Explore Data

```{r}
#| eval: false

## Check out column names
colnames(escape_raw)

## Peak at each column and class
glimpse(escape_raw)

## From when to when
range(escape_raw$sampleDate)

## Which species?
unique(escape_raw$Species)
```

## Clean Data

```{r}
escape <- escape_raw %>% 
    janitor::clean_names()

colnames(escape)
```

## Calculate the Annual escapements by species and sasap regions

```{r}
annual_esc <- escape %>%
    separate(sample_date, c("year", "month", "day"), sep = "-") %>%
    mutate(year = as.numeric(year)) %>%
    group_by(species, sasap_region, year) %>%
    summarize(escapement = sum(daily_count)) %>%
    filter(species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink")) %>%
    ungroup()
    
unique(annual_esc$species)

head(annual_esc)
```

## Using ggplot2

\<!--# Indicate we are using `ggplot()` (call the `ggplot2::ggplot()` function)

1.  What data do we want to plot? (`data = my_data`)

2.  What is my mapping aesthetics? What variables do we want to plot? (define using`aes()` function)

3.  Define the geometry of our plot. This specifies the type of plot we’re making (use `geom_*()` to indicate the type of plot e.g: point, bar, etc.)

**Note** To add layers to our plot, for example, additional geometries/aesthetics and theme elements or any `ggplot` object we use `+`. --\>

```{r}
## All of this code should create the same plot

## Option 1 - data and mapping called in the ggplot() function
ggplot(data = annual_esc,
       aes(x = species, y = escapement)) +
    geom_col()

## Option 2 - data called in ggplot function; mapping called in geom
#ggplot(data = annual_esc) +
#    geom_col(aes(x = species, y = escapement))


## Option 3 - data and mapping called in geom
#ggplot() +
#    geom_col(data = annual_esc,
#            aes(x = species, y = escapement))
```

\<!--# Indicate we are using `ggplot()` (call the `ggplot2::ggplot()` function)

1.  What data do we want to plot? (`data = my_data`)

2.  What is my mapping aesthetics? What variables do we want to plot? (define using`aes()` function)

3.  Define the geometry of our plot. This specifies the type of plot we’re making (use `geom_*()` to indicate the type of plot e.g: point, bar, etc.)

**Note** To add layers to our plot, for example, additional geometries/aesthetics and theme elements or any `ggplot` object we use `+`. --\>

### Boxplot
```{r}
annual_esc %>%
    filter(year == 1974,
          species %in% c("Chum", "Pink")) %>%
    ggplot(aes(x = species, y = escapement)) +
    geom_boxplot()
```

